::: {.callout-important}
## Work in progress
:::

# Introduction

The authors of the present document work as biostatisticians where we
encounter data from many different projects. We have seen many examples
of data files structured in ways that have been demanding for both
ourselves and the final users. Usually, a typical basic course of
biostatistics does not cover the aspects we aim at here, namely how to
set up a good data structure, despite this being a fundamental part of
analysing data.

A PhD student who has only recently begun his or her research career
cannot be expected to know how to structure a data file in a way that
facilitates research on the data for many years to come. Thus, the
student is therefore dependent on their supervisor and others involved
having sufficient knowledge in this area, and also on that knowledge
being effectively communicated early in the doctoral training. This is a
fragile system.

Our hope is that this document will offer useful guidance on how to
avoid issues related to data management, analysis, and archiving,
facilitating work for all researchers/staff involved, both during and
after the course of the research project. We also hope that if the
recommendations are followed in a broader context at the medical faculty
at UMU, this will simplify collaboration across research groups.
Ultimately, the goal is to increase the quality of the research.

It should be noted that the entire document is to be regarded as a "Work in progress". It will evolve, hopefully with input from yourself and other readers.

# What is this document?

This document is intended as a template for how to structure research
data files at the Medical faculty of Umeå University.

The recommendations made here are valid and relevant for projects
especially in their beginning stages. Having well-designed structures
from the outset will usually not increase the total amount of time and
resources spent on data management, but rather simplify subsequent work.

In case you are not involved in medical research, we still hope that the
suggestions made here are relevant to you. After all, even though there
are many fields in science, there are also many common tasks in the
analysis of the data.

We do not claim that the suggestions made here are the only possible
ways to work or the unequivocally "best" ways. We do however claim that
the approach suggested is simple enough to use in most or even all
research projects, and that adherence to the suggestions can be expected
to make your life in research smoother.

# For whom is this document written?

The primary audience is of course those new in research, such as PhD
students. There are however others involved which also could make use of
the recommendations. It is first of all not a given that time spent in
research always results in effective and efficient procedures,
experience is simply not enough. We also hope that the document can be
used by supervisors to ease the way for their students. So the targeted
audience is a general one: staff in any way involved in research - PhD
students, supervisors,... - inexperienced and experienced alike.

# The scope for this document

A very short description of a typical research process, at least the
quantitative type we are focusing on here, is the following figure:


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE) 
```
```{r}
library(DiagrammeR)
mermaid('
graph LR
  A(Raw data, keep unchanged)-->|Cleaning, preferably with scripts|B(Study file)
  B --> |Analysis|C(Results)     
')
```

Eller detta?

```{r, echo=FALSE}
mermaid('
graph LR
  A[Data from others] --> |Import and cleaning|C
  B[Data you manage] --> C[Study file]
  D[Results]
  C -->|Analysis|D     
')
```

In this document we are focusing on the first parts up until "Study
file", *i e* the analysis part is not covered.

There are of course projects where there is only one kind of data
source, either only data from others or only manually entered data, but
the description is meant to cover the following steps:

*   Import of data
*   Cleaning
*   Matching data in more than one file into one single file
*   Manual entry of data

To be more precise, procedures for import, cleaning and matching will
*not* be covered, but rather steps you can take to make the work in all
of them a lot smoother. Manual entry however will be covered.

# For what kind of data is this document written?

Data is stored in different forms and for different purposes. Some forms
are suited for tasks such as *follow-up of individual patients* in their
daily care. Then the use of the file can be simplified by using a
"visual layout", colors, bold text or the like. The goal in such cases
is usually to collect and use data for each individual, not to calculate
summaries such as means for several individuals.

Another purpose is *research* where the goal is to summarize important
aspects of the individual records, and where data on individuals usually
is not important. This latter kind of use most often requires a layout
for the data which is different from the former one. Here we are
concerned with this latter use.

Further, the data considered is of the kind usually referred to as
*quantitative*. Considerations for the structure in qualitative data
must be found elsewhere.

# Why is a good structure in research data so important?

Having a good structure is simply fundamental for working efficiently,
both for yourself during the project and in other contexts.

## Import and cleaning takes time

In the typical statistics course you learn about procedures for
analysis, often with neat and tidy datasets. So when the research on
real data is begun, many less experienced researchers are surprised or
even overwhelmed by the amount of time needed for importing data files
and cleaning the data. Further, if the documentation for variables and
codes is lacking, the analysis takes unnecessarily much time simply
because you have to go back to other sources in order to find out what
you are working with.

## Errors

There are also often flaws and errors in the data. You can hope to find
them early in the cleaning process, but even if they turn up later, a
good structure simplifies finding and correcting these errors.

## Projects often live long

Many research projects have an afterlife. After some years, you might
yourself want to do a follow-up study: what are the long-term effects?
There could be other researchers interested to make such a followup. A
supervisor might want to put previously collected data in the hands of
new PhD students. You could be contacted by other researchers who are
interested in your work, ending in a new joint project. In all those and
many other situations there is a definite need for the original datasets
to be well-structured and well-documented.

## Cooperation

Some projects are from the start set up as based on more than one
researcher or research group. Data can be collected in multiple centres,
and the combined dataset can also be used in multiple centres. In such
cases, there is a definite need for well-designed data structures.

## Archiving/transparency

When research projects have reached their final days, many just wither
away. Given the resources spent, this seems as a poor outcome. It is
both a matter of making it possible to wake projects for future research
and part of making the research efforts transparent. The latter could
simply be seen as an important part of maintaining the publics' trust in
the scientific community. So the research data should eventually be
stored and archived in a manner that makes it possible to dig up even
previously buried projects.

## Reproducibility/open data/open science

There has been an increasing emphasis in many areas on
*reproducibility*, on *open data* and *open science*. None of these
goals can be attained if the underlying datasets are poorly documented,
stored in formats no one can access and the like. However, there are
limitations to what can be made open when it comes to data in typical
medical research. It is also an area where both practical and legal
issues have to be considered, so it is often necessary to consult legal
and technical staff.

# What aspects are *not* covered in this document?

## The structure of data files and folders

A structured approach to storage of data and other documents in a
research project is of course important. However, the level in this
document is on a somewhat lower level with a focus on more practical
aspects of data management. A good source for more details on structures
on the higher level is an 
[article](https://lakartidningen.se/klinik-och-vetenskap-1/2013/02/att-strukturera-och-dokumentera-forskningsprojekt/) by Eloranta et al, as well as information at the [UMU library
web](https://www.umu.se/bibliotek/forskningsdata/samla-och-organisera/).

## Anonymous or pseudonymized data

In most research involving humans, issues of integrity and handling data
on individuals securely are important. At the same time, the subjects
are unique individuals, and some way of keeping track of everyone is
needed. It has become common to use some *key*, an identifying variable.

Procedures and legal frameworks for dealing with these issues are mostly
general, but the local implementation is often specific to the actual
organization. So in this document the choice has been not to cover these
issues but rather to refer to frameworks which should be at place where
you work.

## FAIR data

The concept of FAIR data (Findable, Accessible, Interoperable and
Reusable) has been discussed during the last decade. As the [GO
FAIR](https://www.go-fair.org) network describes the ambitions: 

> The ultimate goal of FAIR is to optimise the reuse of data. To achieve this,
metadata and data should be well-described so that they can be
replicated and/or combined in different settings.

The focus is on systems and structures to make data available and is
part of a "movement" emphasizing "open data" and open science". This
implies a focus on both legal and practical arrangements - repositories,
identifiers for data,... - and are as such commendable. The suggestions
put forward in this document certainly are in line with the suggestions
made by the GO FAIR network. However, we limit the scope for this
document to more direct suggestions for how you should work with your
own datasets to simplify your research life. The possible extension into
making research data comply to the FAIR principles is covered by
university libraries, see for example [UMU: Research
data](https://www.umu.se/bibliotek/forskningsdata). The recommendations
made here should make this process smoother.

## Storage of data

Researchers and students involved in the kind of research we have
focused upon here are active in different organizational settings:
mainly within universities, mainly in clinical work or mixtures of the
two, within a smaller group or with a multicentre approach spanning an
entire country or even several countries. Depending on the organization
you find yourself in, there are hopefully structures in place for
storing your research data in a secure way. However, the practical
arrangements are so different that it is not possible to provide any
specific suggestions here.

## Data management plans

The concept of data management plans (DMP) has become emphasized in many
contexts. Funding agencies increasingly require that applications
contain such DMP's. The basic idea is that a DMP should provide
information on how research data is managed throughout the project, and
the purpose is to ensure that data is handled securely, legally and
ethically. A good source is [Checklist for data management
plans](https://snd.se/en/resources/checklist-data-management-plans)
provided by [the Swedish National Data Service
(SND)](https://snd.se/en).

## Archiving

There are legal requirements for the archiving of completed research
projects: *what*, *how*, *for how long* are typical questions.
Universities and health care providers have their own implementations
for the procedures involved.

## Manuals for particular software systems

Software used for statistical analyses encompass a range of systems,
from spreadsheets to more comprehensive systems specifically designed
for statistics. There is a lot of material for almost any software, such
as style guides or more technical details for how the system should be
used. The scope for this document is however a general one, so the main
text will not cover details for any particular system. There are however
some software specific notes for some of the aspects covered in the main
text in the [Appendices](#appendices).

# Principles for data file organization

The most important advice is to be as *consistent* as possible
irrespective of the chosen structure.

## Examples of common problems

The list of problems when dealing with research data can be made very
long. There are however some more common problems: 

* Inconsistencies in naming variables can make coding and writing challenging 

* Lack of documention of variable content and codes used can make calculations or
creation of new derived variables more or less impossible 

* Data structured in visually appealing format - parts of data placed in
different spots in the file, colored cells,... - is often difficult to
process for statistical analysis 

* A database is often a changing entity. A structure which is not well-defined in advance can make the
subsequent work a lot more difficult than needed. 

* New data points might be added, others can be filtered out from analysis. 

* New derived variables are created during the project. 

* Dates can be stored in a number of formats, and the process to convert dates into a format that
works in your software can be hard. 

* Exchange of data between different software systems can be difficult. 

* Datasets collected in more than one centre are often marred by inconsistencies in the formats used by the
centres involved, making the compilation of data challenging.

## Definitions

| Concept/term | Intended meaning |
|------------------------------------|------------------------------------|
| `Individual` | Any kind of "basic unit": a person, an animal, an organization,... It does not necessarily denote a human, but the term *individual* will for the of simplicity be used throughout this document |
| `Observation` | Often called *record* or *case*, where there could be one observation or more for each individual |
| ? | [Andra begrepp som bör definieras?]{.mark} |

## Basic structure of the study file

The basic structure of your study file should typically be *one*
rectangle, preferably as a complete rectangle with no gaps, see [Missing
values](#missing-values). This however can be achieved in a number of
ways. There are cases where you have multiple observations for each
individual, and here the rectangle can be set up with separate variables
for the different time points, a so called *wide* format, or with
multiple rows for each individual, a *long* format. Depending on the
calculations you want to make, it could be necessary to use either of
these setups, or even both, see [Wide or long
format](#wide-or-long-format). In both cases, you have a rectangular
design.

It should also be noted that data often comes in separate files where
each file preferably should be organized as such rectangles. Having such
multiple files could be the best way to store and archive your data, but
for the analysis you want to do, the relevant pieces usually have to be
merged into one single rectangular study file, see [Compilation of data
split into pieces...](#compilation-of-data-split-into-pieces-into-one-dataset).

## Tidy data

There are many ways to organize data where subsequent processing can be
difficult. The ways in which they should be organized to make processing
easier is however not that many. Some general recommendations: 

* Let each column correspond to one variable 
* Let each row correspond to one observation 
* Let all variables in the dataset have the same
unit of observation 
    [Den här skrivningen känns svår. Räcker det med att skriva
det som kommer efter? förklara vad som menas. Handlar detta om att data
kan organiseras i flera "tabeller", ex en med "basic charcteristics" och
en rad per skalle, och en annan med flera rader per skalle? Det är i
alla fall ett av Wickhams förslag.]{.mark} 
* Let each cell contain only a single value

In the following examples, there are often combinations of flaws, not
just the ones mentioned in the header. Especially important are the
missing cells, see [Missing values](#missing-values).

## Each column corresponds to one variable in the dataset

Bad example

id    female   male   age
---   -------  -----  ----
1     yes             34
2              yes    38
3              yes    42

Another bad example

id   stage_1    stage_2   stage_3    stage_4 
---  --------   --------  --------   --------
1    yes                                
2                         yes              
3               yes                       

A good example

id   sex   age
---  ----  ----
1    f     34
2    m     38
3    m     42

Another good example

id   stage 
---  ------
1    1    
2    3    
3    2    

When the dategories are mutually exclusive, there is only need for one
variable/column. When there can be more than one category, as when the
question is "Which of the following medications have been prescribed?",
this could be dealt with in several ways, see [Multiple
response](#multiple-response).

## Each row corresponds to one observation

Bad example

id    variable   value  
---   ---------  -------
1     sex        female 
1     age        34     
2     sex        male   
2     age        38     
3     sex        male   
3     age        42     

A better example

id   sex    age 
---  ----   ----
1    f      34  
2    m      38  
3    m      42  

Another bad example

id    time   sex   age   hba1c 
---   -----  ----  ----  ------
1     1      f     34    32    
      2                  29    
2     1      m     38    33    
      2                  30    
3     1      m     38    31    
      2                  32    

A better example

id   time   sex   age   hba1c 
---  -----  ----  ----  ------------
1    1      f     34    32    
1    2      f     34    29    
2    1      m     38    33    
2    2      m     38    30    
3    1      m     38    31    
3    2      m     38    32    

If the unit of observation is a unique combination of more than one
variable, such as here with `id` and  `time`, then all cells should be filled
with the correct content.

## All cells contain only one value

Bad example

Here the variable `birthplace` is meant to cover geographical region *and*
municipality:

id   sex   age   birthplace
---  ----  ----  ----------- 
1    f     34    24A 
2    m     38    25C 
3    m     42    24B

Good example

Use separate variables for region and municipality: 

id   sex   age   birth_region   birth_municipality
---  ----  ----  -------------  ------------------- 
1    f     34    24             A 
2    m     38    25             C 
3    m     42    24             B

# Variable names

We recommend using a simple structure as it well works everywhere. Be
consistent.

## Use names in English

*   Do not use diacritics such as *å*, *ä*, *ö* or the like in variable
    names

Bad example: `Åtgärd`

Good example: `treatment`

## Separate parts in names with an underscore ("\_")

*   Do not use spaces, dots (".") or upper case letters to separate
    parts

Bad example: `Type of primary cardiovascular event`

Better example: `type_of_primary_cardiovascular_event`

An even better example: `cv_type`

Use of underscore is often called "snake_case".

## Keep your variable names short

Name variables so that the names provide a sufficiently clear
description of the content. They do not have to be exhaustive. Short and
meaningful is worth striving for.

Bad example: `type_of_primary_cardiovascular_event`

Good example: `cv_type`

## Use only lower case letters

*   Do not begin variable names with upper case letters
*   Do not use names with mixed cases ("CamelCase")

Bad example: `RespondentStatus`

Good example: `respondent_status`

## Do not include explanations of codes in the variable name

Bad example: `diabetes_1_yes_0_no`

An even worse example:
`Type of primary event: 1= IS; 2=ICH; 3=SE; 4: Bleeding BARC 3a,b,c or 5; 5: All cause of death`

Good example: `diabetes`

Another good example: `type_primary_event`

# Documentation of variables and codes

## If possible, assign explanatory "labels" to variables

Variables should of course be understandable. This is however not always
that simple to achieve, especially if a variable name is short and there
are many versions of the content, *e g* `dog1` and `dog5` for a child
being allergic (yes/no) to dogs at 1 and 5 years of age respectively. In
such cases, it is critical to assign good explanations to the variables.
How this is done depends on your software. Irrespective of how you set
up such explanatory labels, you should always keep a system for this
kind of information.

If the software you use allows such labels to be set, then your
analyses could also be simplified with such labels since the output -
tables and graphs - in many software systems will make direct use of 
these labels in full text.

## Categorical variables - numerical or text/string?

Categorical variables such as age group *can* consist of text in string
formats, *i e* stored as text rather than numbers, such as "25-64", "65-".
Most statistical software can handle such formats. However, it is often
simpler to use numerical codes, so if you use string formatted
variables, be prepared to convert these to numerical ones.

## Codes should be assigned explanations

Categorical variables consist of markers for categories, such as age
group. The codes used in such variables must be assigned explanations.
The suggestions made for variables are valid or codes too: they should
be informative but not longer than needed. These codes are an important
part of your [codebook](#create-and-maintain-a-codebook).

If you work with spreadsheets, you could at least use comments.

## Variables measured at different time points {#variables-measured-at-different-time-points}

Given that the dataset is in a wide format (see [Wide or long
format?](#wide-or-long-format), variables measured at different time
points should be given exactly the same name, except for a (short)
suffix.

Bad example: `Hba1c_at_baseline`, `hba1c_at_3month_followup`,
`hba1c_12_mon_fu`

Good example: `hba1c_0`, `hba1c_3m`, `hba1c_12m`

If you have a [long](#wide-or-long-format) format for this kind of data,
then the `hba1c` data should consist of one variable with the actual
`hba1c` levels and another (`time`) with values such as 0, 3, 12 for
each observation.

## Create and maintain a codebook {#create-and-maintain-a-codebook}

A codebook describes the contents, structure, and layout of a data
collection. More specifically explanations of variables, units for
measurement variables, explanation of codes used in categorical
variables, codes used for missing values and other attributes of the
variables in your dataset. There are routines in the software most
commonly used to generate such codebooks, but it could also consist of
an Excel file with one sheet containing explanations for variable names,
another sheet with explanations of the codes used. A short example of a
Excel code book is found in one of the appendices.

# Wide or long format? {#wide-or-long-format}

Many datasets consists of data points from one measurement period, and
the basic structure can be a fairly simple "rectangular" set of data. It
is however also common with datasets where *time* is a fundamental
variable. Follow-up of patients makes it necessary to have data on
patients from baseline, from 3 months, 12 monts and possibly later than
that. The structure in such cases has to take these different times into
account, and there are basically two ways to do so: *wide* and *long*
formats.

There is no single correct structure for such data, it can even be
necessary to have both structures in parallel due to the details for
your software. As a consequence, switching from one format to the other
is a common task. It can however be quite demanding, so it could be wise
to consult a statistician when the need arises.

Naturally this also applies for other cases where there is some
repetition of similar measurements: different places, different parts of
the body etc.

Data from such instances in time can also be stored in different ways: 

* As one file for each point in time 
* As different worksheets in the same spreadsheet 
* With all the observations in one single file

If the structure is one of the two first you will probably have to join
the parts together sooner or later. Such joining is described in [Data
on the same individuals...](#data-on-the-same-individuals-in-more-than-one-place).

## Wide format

Here the basic observation is one "individual": one row per individual,
several variables for each timepoint.

Example:

id   sex   hba1c_1   hba1c_2   hba1c_3 
---  ----  --------  --------  --------
1    m     34        36        38      
2    m     32        32        33      

## Long format

The basic observation in such a structure is the combination of
individual *and* timepoint. Then there is one row for each such
combination, resulting in several rows for each individual.

Example:

id   sex   hba1c   time 
---  ----  ------  -----
1    m     34      1    
1    m     36      2    
1    m     38      3    
2    f     32      1    
2    f     32      2    
2    f     33      3    

# Missing values {#missing-values}

Few datasets are free from missing values, and there can be many reasons
behind: 

* Survey respondents might have left questions unanswered. 

* In a study of the effects of some treatment, participants can drop out from
planned followup. 

* Medical records often have gaps for variables which
at the time for registration were not considered as sufficiently
important to register.

Such gaps should be treated with care. A general advice is to put effort
into keeping track of such gaps since the subsequent analyses could be
hampered. Some typical problems:

*   In order to make readers of your results trust your findings, you
    should be able to keep track of and report such reasons and the
    frequence with which they have occurred.
*   Depending on software, missing values are in some systems coded with
    numerical values, such as 999 eller -999. If these are not defined
    as missing values, any calculation of such a variable will be
    erroneus.

Generally, it is good to deal with missing values explicitly so that you
can distinguish between truly missing values and unintentionally missing
values. In other words, it is recommended to work with files where gaps
in your original data have been filled. This is especially important in
the raw data file. If you find empty cells later on in the process, you
know that their origin lies in the calculations you have made.

An approach which works in all systems is to accompany a variable with
missing values with another separate variable/column where the reasons
for the missingness is recorded. This could however result in a large
amount of variables. There are also other built-in mechanisms to deal
with missing data in different software systems, see notes for some 
specific systems in the [Appendices](#appendices).

You should also scan through your data and codebooks accompanying your
data and look for indications of how missing values have been dealt
with. A general advice, especially when you work with datasets delivered
to you, is that you in an early stage go through the variables. Go
through the data in any way you can imagine, e g by sorting (both
ascending and descending order) to help you find missing values. You
need to make your self acquainted with your data.

# Miscellaneous

## Data on the same individuals in more than one place {#data-on-the-same-individuals-in-more-than-one-place}

It is sometimes good to store data for the same individuals in different
files. However, those files must often be combined into one single study
file before you can begin with your analyses. Then it is necessary to
have a common "key" variable in all parts meant to be joined:
"personnummer" or some other identifying variable. The "key" could also
consist of more than one variable, e g `id` *and* `time` if the basic
unit of observation is one particular individual at a specific point in
time.

For variables with any kind of repetition, in time or otherwise, it is
also good to have the same names and formats when data in parts are to
be joined, see [Variables measured at different time
points](#variables-measured-at-different-time-points).

## Compilation of data split into pieces into one dataset {#compilation-of-data-split-into-pieces-into-one-dataset}

Data stored in separate parts must in most cases be compiled into one
dataset, a process often called "matching", "merging" or "joining". The
most straightforward matching is when you have one row for each
individual in all files, a 1-to-1 matching. Another kind of matching is
when you want to assign information from one file with unique records
for each individual into a file with many rows for the individuals
involved, a so called 1-to-n or 1-to-many matching.

These matching procedures can be made in any decent statistical
software. It could however be quite demanding to make it work, so do not
hesitate to consult a statistician.

## Import of files from others

It is common to have datasets coming from other sources:
Socialstyrelsen, other data bases and the like, where their export is
done into files in formats you do not use yourself. For example are
files from Socialstyrelsen often SAS files, and you might work wth SPSS,
Stata or R. Files of this kind can almost always be imported directly
into your statistical software.

There can still be problems depending on the structure in the files you
receive. One common example is when a numeric variables such as dates
are formatted as "string" variables rather than numeric variables. Such
variables can usually be converted into the format you need, but the
time needed for such conversions can be quite demanding.

As a general rule, try to communicate with the organization, preferably
*before* they send files to you, in order to see if they can send files
in formats which are as easy for you to work with as possible. Do not
simply accept files in the format they use.

Another advice is to see if you can get the source to send files in both
*their* preferred format and other formats. It might be possible for
them to send SPSS, Stata or R data files. They can also always export
data files into non-proprietary formats such as comma or tab-delimited
files. You can always import such files yourself. In this process, you
are usually able to control the finer details of the import, so even it
this could be seen as a rather laborious workaround, it can still take
less time than converting variables *after* import.

## Should all parts of the datasets be put in good order?

As has been stated elsewhere in this document, data often comes from
different sources: your own manually entered data, external databases,
data collected by others or the like.

In the case of data from other sources, you are not in complete control
yourself. There might be a lot of variables already named, having more
or less complete documentation, and the structures might not be in line
with the recommendations made here. Then you have to make a decision
about to what extent "old" structures should be changed. There is not
any simple rule for how to proceed in such cases. You could limit your
efforts to your own nwewly derived variables only, or you could set out
to change all of the old variables.

Our recommendation is that whatever approach is chosen, you should try
to make these decisions early in the process. It does not become easier
to set up a good structure for the data later.

## The raw data file

The original datafiles containg raw data should be kept raw, *i e*
unchanged, preferably also in some other folder than cleaned files
and/or files with new derived variables. Raw data files could also be
write-protected as an extra countermeasure against possible errors.

## How should additions and corrections to data files be made?

It is easier to say how additions and changes should *not* be made:
manual changes in the datafiles. Such changes seldom leave any trace,
and it could become impossible to to understand what changes have been
made. A better approach is to use scripts/syntax. If scripts are not
used, then all such changes should be clearly documented in a log book.

Changes should also be made in new copies of the file rather than
keeping one single file, see [Version control](#version-control).

## Backup
Backup is important. The simple solution is to store all files at secure 
servers made available where you work. If such resources are not available 
or if you for some other reason want to set up backup procedures yourself, 
make sure backup is done systematically and with regular intervals. 

## Version control {#version-control}

There are systems available for version control, e g "git" or "github",
which are particularly suitable for R users. However, for users of other
systems, git may not provide a smooth workflow, so other alternatives
should be used. A simple solution is to add the current date in the file
names whenever a change is made. In that way, you might perhaps end up
with many different but similar versions of the files. However, that is
a small problem, and one that you can solve in the later stages of the
process when there are files you clearly do not need anymore.

## Multiple response {#multiple-response}

One fairly common type of data which needs extra attention is when there
could be more than one answer, as for example to the question *Which of
the following medicines have been prescribed during the last year? One
or more alternatives can be marked* followed by possible alternatives
such as "substance A", "substance B" and so on.

This kind of data can be dealt with in more than one way:

### Only a few combinations, mostly only one of the alternative is chosen

One single variable, combinations are entered as more than one code: 

id   medicine 
---  --------- 
1    a 
2    a 
3    b 
4    ab 
5    c

This is technically possible, but not recommended. One common
consequence is that there could be many different combinations, making
it difficult to effectively summarize the variable..

### Many possible combinations

One column/variable for each alternative, where each alternative is a
binary variable either with 0's and 1's or simply 1's: 

id   med_1   med_2   med_3 
---  ------  ------  ------ 
1    1       0       0 
2    1       0       0
3    0       1       0 
4    1       1       0 
5    0       0       1

This is usually the best approach, both because the data can easily be
summarized and because all information is available and fairly
strightforward to set up into other formats.

# Advice for manual entry of data

The recommendations made in this document are meant to be valid for any
kind of data, irrespective of whether the source is someone else or if
you have entered data manually yourself. Such manual entry is an
important stage where you should strive to minimize the risk for errors
and to make the entry process as smooth as possible.

## Systems for data entry

Data could be entered in a number of ways:

System                                            Comments 
-------------------                               -----------------------------------------------------------------
Excel/spreadsheets                                Possible, flexible and available, but not particlarly safe. Consider adding "validation" to the data entry. 
Statistical software (SPSS, Stata R,...)          Works similarly to spreadsheets, perhaps less flexible and usually without validation procedures. 
Redcap or other systems for "case report forms"   Works quite well, validation is possible, data can be exported to formats suitable for import into statistical software. 
Database systems: MS Access, SQL,...              Perhaps the safest tool, however setting up a good database structure is not always that straightforward. 

## Keep the structure simple in order to minimize errors

Whatever structure, one single file for all data or data split in
separate parts, make your utmost to set up a *simple* structure which is
easy for all involved parties to understand and follow. Simplicity,
consistency and ease of use is the best way to minimize errors and to
avoid timeconsuming pitfalls.

## Different worksheets or different files for data in the same individuals

If you use a spreadsheet for manual entry and there are different
sources for the basic sources, such as repeated data from different time
points, then the different pieces can be stored as separate files or as
separate worksheets in one single spreadsheet.

## If there are questions, put comments in adjoining columns

Bad example: 

id   sex   hba1c_1   hba1c_2   hba1c_3
---  ----  --------  --------  --------
1    m     34        36?       38 
2    m     32        32        33 (error?)

Good example: 

id   sex   hba1c_1   hba1c_1_c   hba1c_2   hba1c_2_c   hba1c_3    hba1c_3_c
---  ----  --------  ----------  --------  ----------  ---------  ---------- 
1    m     34                    36        ?           38         
2    m     32                    32                    33         error?

## Dates

In many research projects, *time* is involved. Counting the time from
diagnosis to death and delimiting the records to some specific time
period are common tasks. In order to make the treatment of time
variables as simple as possible, there are some general suggestions
especially relevant when you enter dates manually:

*   Be consistent. Whatever way dates are entered, use one system.
*   Avoid formats for dates which can be misunderstood: 121110 could
    mean Nov 10 2012, Nov 12 2010 or even Dec 11 2010.
*   A good way to enter dates in line with international standards (ISO
    8601) is the YYYY-MM-DD format, *i e* four digits for the year and two
    for month and day respectively. The "-" separator is preferred, but
    not always necessary.
*   A safe way to enter dates (and times if needed) is to use separate
    variabels for year, month and day (and possibly hour, minute and
    second too). There are most certainly procedures in your statistical
    software which can be used to combine these into a working date/time
    variable.

## No calculated values in the raw data files

[Calculations, avser det formler i Excel-blad? Eller calculated/aggregated values, typ medelvärden, standardavvikelser/...? Eller färgade celler?]{.mark}

[(Ett försök)]{.mark}

Bad example: 

id    sex   hba1c_1   hba1c_1_c   hba1c_2   hba1c_2_c   hba1c_3   hba1c_3_c
----  ----  --------  ----------  --------  ----------  --------  ---------- 
1     m     34                    36        ?           38         
2     m     32                    32                    33        error?
.     .     .         .           .         .           .         .
.     .     .         .           .         .           .         .
.     .     .         .           .         .           .         .
356   f     31                    31                    30    
Mean        32.3                  32.1                  31.9    
Std         3.4                   3.5                   3.6 

Such additions of means and/or other calculated entities have no place
in your raw data. In the best case, it only makes the structure more
complicated than necessary. In the worst case, you end up using such
aggregated values in your subsequent calculations.

## Multicentre studies

Multicentre studies are prevalent. In most cases, there is an overall
ambition to collect data in the centres in a streamlined fashion. Our
experience however is that the resulting parts from the centres often
are unnecessarily different. Some examples:

*   Formats for dates differ, one centre could have used "2025.01.01"
    and another "250101"
*   In one centre, someone finds ways to "simplify" their own data entry
    using special formats, not taking into account that their part then
    becomes different from the parts from other centres
*   Data is structured and stored in an Excel file with a "visual"
    layout rather than having separate files or at least separate
    worksheets for each centre
*   ... (Andra exempel?)

The overall recommendation is that you devote time and energy on
communication and calibrating the procedures for data entry. Also try to
make all involved understand that even small changes in the data entry
procedures will cause problems. On a more technical level, if you use
Excel or similar, try also to add "validation" of the contents in the
columns in order to minimize the risk for errors and inconsistencies.


# Consistency again

It is often not so easy to be consistent. During the research process
where you assign namns to new derived variables, add explanatory labels
to the variables and to codes used, maintain a codebook, there are a
number of steps where choices have to be made and where inconsistencies
can enter. As authors of this document, we cannot claim that we succeed
in being entirely consistent ourselves. The basic message here is
however that you should strive to be as consistent as possible. The more
consistent, the better. And you should not hope that it will get any
easier in the future. The gaps left behind today will most often not be
covered later.


# Syntax/scripts/do files {#syntax-files}

We strongly recommend that you work with "syntax" as a way to run your
processing of data, *i e* setting up "code" which is stored in specific
files for whatever you want to achieve, code which subsequently can be
executed. It is not uncommon that users begin with using menus rather
than syntax, so it could be preceived as a hurdle. However, our
experience that there are many advantages with using syntax, and not
that many disadvantages:

*   You can add explanatory comments in the syntax files, making it much
    easier afterwards to see what you have done.
*   A syntax files is a rather complete record of what you have done.
*   If data is changed, e g additions of new records or the deletion of
    others, then going through all steps already done through the use of
    menus is often not possible or at least very demanding. With syntax
    it is often a matter of a few clicks.
*   In many statistical procedures there are options which might not be
    available in the menus.

Depending on the software, the creation of syntax can be more or less
straightforward, but it can always be done. Put in another way, one
reason, along with several others, why we would not recommend use of
spreadsheets for more systematic statistical work is that there is not
any simple way to run such software with syntax.

When you choose to use syntax, a basic recommendation is to have different scripts for different tasks,
typically

* Import of data file 
* Data cleaning 
* Analysis, possibly in steps.

You could of course use a single syntax file for all parts - import of files, cleaning, analysis and presentation - but this tends to result in long files which can be difficult to navigate your way through. 

# When the project is about to be closed?

*   Contact the person or unit responsible for archiving in order to
    find out what parts of your files and documents should be archived,
    what formats they prefer and how the whole package should be stored.
*   Create codebooks for your data files.
*   Save the codebook in different formats: as plain text, pdf, HTML,
    Word, Excel, simply in any format which could be suitable for
    another reader.
*   Save the data in the format used in your software system.
*   Save the data in non-proprietary formats such as CSV or tab
    separated, perhaps also in proprietary formats such as Excel or SAS
    files. Here the most important are the non-proprietary formats since
    such files should be possible to read in all systems. However,
    keeping other formats can simplify future use.

# Some final warnings

*   Don't calculate the time between events by manual calculations based
    on your calendar. There are always better ways to do this in your
    software.
*   Don't try to combine data stored in separate files by manual
    cut-and-paste operations. See [Compilation of data split into pieces
    into one dataset](#compilation-of-data-split-into-pieces-into-one-dataset).
*   Don't calculate new variables for combinations of your basic
    variables, such as a binary variable for "males_age_34_and_older",
    unless it is absolutely necessary. In most cases, you can use the
    two variables "gender" and "age" together to filter out such
    records, leaving you with a lot fewer variables.

# Other support at UMU

Most or even all university libraries provide support for working with research 
data. At UMU, you can read more at [Hantera forskningsdata](https://www.umu.se/forskare/planera-och-genomfora/hantera-forskningsdata).

# References


*   Broman, K. W., & Woo, K. H. (2018). Data Organization in
    Spreadsheets. The American Statistician, 72(1), 2–10.
    <https://doi.org/10.1080/00031305.2017.1375989>
* Eloranta, S. et al, (2013). [Att strukturera och dokumentera forskningsprojekt](https://lakartidningen.se/klinik-och-vetenskap-1/2013/02/att-strukturera-och-dokumentera-forskningsprojekt). Läkartidningen, nr 8.    
*   Wickham, H. (2014). Tidy Data. Journal of Statistical Software,
    59(10), 1–23. <https://doi.org/10.18637/jss.v059.i10>

# Other sources

*   [The FAIR principles](https://www.go-fair.org/fair-principles)
*   Svensk Nationell datatjänst (SND), <https://snd.se/sv>
*   Reproducible research?
*   [R for Data Science (2nd ed)](https://r4ds.hadley.nz/)
*   Corti, L., Van den Eynden, V., Bishop, L., Woollard, M. (2019),
    [Managing and Sharing Research Data - A Guide to Good
    Practice (Second Edition)](https://uk.sagepub.com/en-gb/eur/managing-and-sharing-research-data/book262873), SAGE
*   Collier, J. (2009), [Using SPSS Syntax: A Beginner's
Guide](https://uk.sagepub.com/en-gb/eur/using-spss-syntax/book229167?_gl=1%2A1emm9pb%2A_up%2AMQ..%2A_ga%2AOTI1NTI3NDMuMTc4MDM5OTgyMQ..%2A_ga_60R758KFDG%2AczE3ODAzOTk4MjAkbzEkZzAkdDE3ODAzOTk4MjAkajYwJGwwJGgxMjMxNzc2MzU0), SAGE
* Long, J.S. (2009), [The Workflow of Data Analysis Using Stata](https://www.stata.com/bookstore/workflow-data-analysis-stata), Stata Press, Stata Press

# Appendices {#appendices}

## Codebook in Excel

[Min (RL) tanke här är att ge förslag på hur filinformation ska sparas ner om man nu nödvändigtvis vill ha det i Excel, eller kanske allra helst csv-format. Typ:

* En flik/fil med grundläggande information om datamaterialet: projektnamn, år, ansvarig,... Eventuellt referenser till metodbeskrivningar?

* En flik/fil med variabelnamn och variabelförklaringar, helst i separata kolumner

* En flik/fil med variabelnamn och kodförklaringar, gärna med en rad per variabel och flera kolumner för koder och förklaringar, typ “sex|1|man|2|kvinna|” (utan |). Den strukturen är rätt lätt att importera till SPSS och generera syntax för ```VALUE
LABELS```. Samtidigt är den hyfsat generell och någorlunda lättläst.

* Annan information, ex enheter (“mmol/ml”), helst med variabelnamn och enhet (eller annat) som separata kolumner. Detta kan vara bra för arkiveringsvinkeln också i och med att det kan sparas ner i ett icke-proprietärt filformat. Eller är detta överflödigt och/eller kan lämnas till UB:s förslag/mallar?]{.mark}


## Style guides

There are many style guides for the use of specific software, such as
[Google's R style guide](https://google.github.io/styleguide/Rguide.html#:~:text=R%20is%20a%20high-level%20programming%20language%20used%20primarily,the%20Tidyverse%20Style%20Guide%20by%20Hadley%20Wickham%20license)
or [The DIME Analytics Stata Style
Guide](https://worldbank.github.io/dime-data-handbook/coding.html#the-dime-analytics-stata-style-guide).

It is however important to realize that there are many such guides, and
they might not be telling the same story.

## Notes on R

### Missing values

## Notes on SAS

### Missing values

## Notes on SPSS

### Logging steps taken

One way to get an explicit account of the steps taken is to add the
syntax text to the output generated by the software. Go to
`Edit/Options` where you go to the `Viewer` window. There the "Display
commands in the log" can be marked. Besides giving a more complete
account of what you have done, the text can also be copied into syntax
files.

There is also a "journal" keeping track of what has been done, which is
simply a text file with the underlying syntax. This journal can be used
as a security measure in case you have not remembered to save files as
intended. The default location of the journal file can be seen in the
`File locations` part of the `Edit/Options`.

### Variable attributes

In SPSS, variables have *attributes* such as `type` (numeric, string,
date etc), `missing`, `measure` (nominal, ordinal, scale) and some other
aspects of the variables at hand. The `measure` attribute is
particularly important since calculations or representations of your
variables are in many routines determined by this attribute. Two
examples are *Custom Tables* and *Chart Builder*, where for a scale
variable the default is to calculate a mean, whereas the default for a
nominal variable is a count. As a general rule, it is recommended that
attributes are set for your variables of interest. However, in those
routines where the measure attribute is important, you can make
temporary changes of the measure level.

You can also define your own attributes, which could be a way to store
other types of metadata, units for measurements is one simple example.
In the menu system, the place to define such attributes is
`Data/Define Variable Properties`. These attributes can also be included
in the codebook.

### Missing values

Some sources recommend that missing values in numeric variables should
not be coded with numeric values, but rather by other means. This does
however *not* work in SPSS. A numeric variable can only contain numeric
values, so any attempt to use both numbers and string data in the same
variable results in it being defined as a string variable.

The way missing values for numeric variables is treated in SPSS is to
use the *system missing* and *user defined missing* status. *System
missing* is simply an empty cell, displayed as a dot (.) in the cell.
*User defined missing* means that you can define a specific number as a
missing value, which makes the system exclude that cell in all
calculations. You can also set more than one value to discern between
different reasons for the missing mechanism, e g "77: no answer", "88:
wrong answer", "99: no known reason for missing" or the like. This also
requires that labels are assigned to these different values.

It is generally common to use "impossible" values to define user
missing, e g -9 in a variable for height, or 999 in a variable
containing answers to a question with five alternatives in a
questionnaire.

The use of numerical values to define missing values makes subsequent
processing of the data relatively simple, but there is also a definite
drawback: all values intended to be markers for missing values *must* be
defined as such. In case you get datasets from other sources, take your
time to check if such values have been used. If you do not set the
missing defintions correctly, all calculations based on variables with
these gaps will be erroneous. In case you use procedures for imputation
of missing values, then these procedures will also not work properly if
there are values which should be defined as missing values but are not.

Besides using syntax for the definition of missing values, you can also
use `Data/Define Variable Properties` in the menu system. The missing
definition can also be set in the `Variable View` window, but since
changes there are done without trace, this is not the recommended way.

For string variables, there is no specific code for a missing value, it
is simply shown as an empty cell. You *can* define missing values for
string variables, but it is not that common. A more common approach is to
convert categorical string variables into numeric ones (see e g
`Transform/Automatic Recode`) followed by defining numerical values as
user missing.

Another issue is the way SPSS treat missing values in assignments: 

* Use of operators such as "+" for addition yield missing values when there
are missing values: `COMPUTE sum=var1+var2.`gives a missing a value if
one or both of `var1` and `var2` are missing. 
* Use of functions such as `SUM` uses all non-missing values: `COMPUTE sum=SUM(var1, var2).` sums
the content of `var1` and `var2` when one or both them are non-missing.

### Codebook

Codebooks can be produced by use of `Analyze/Reports/Codebook`. The
result can preferably be placed as the single output in an output window
which subsequently can be saved as an SPSS output file, but perhaps even
better as a pdf or HTML file.

The structure in the codebook is also determined by the attributes set
for your variables, most importantly the measure attribute. In order to
make the best documentation of the data, it is clearly recommended that
you assign "correct" attributes to your variables as fully as possible.

### Syntax files

The overall recommendation is that you use syntax for your processing.
With time, you might learn how to write syntax directly for some
commands. However, the easiest tool is usually the `Paste` button found
in almost everywhere. So press `Paste` instead of `OK`. Often directly,
sometimes after you have experimented with different versions by use of
menus and the OK button, but after which you press the Paste button when
you think you have come up with a more definite version.

The syntax files can also become quite large. As a way to split the
commands into more manageable chunks, see [Syntax/scripts/do files](#syntax-files), 
you can use a "master" syntax file to invoke sub files by `INSERT FILE` (for details,
see the help documentation for `INSERT`). Another way is to have one
file for reading the raw files, another for the cleaning, in which you
invoke the previous file in the beginning with the `INSERT FILE`
command.

### Help

Help files can be found for almost every part of the menu system. One
way to find such support is to use the `Help` button found in most of
the windows/frames opened through the menus. For help with syntax,
pressing F1 will open the help file for the actual command your marker
is placed in. You can also see the `Help` menu in the menu system.

## Stata

### Missing values
